---
editor_options:
  markdown:
    wrap: 72
output: pdf_document
runtime: shiny
---

# Project 2: Visualization of global precipitation and global temperature

Team C

Nguyen Nhat Minh (V202100523)

Hoang Trung Thanh (V202100516)

Tran Anh Vu (V202100569)

## Introduction

Our main questions are:

-   What is the difference between temperature and precipitation between
    regions in the world ?
-   Is the change in surface temperature and precipitation in each
    region over time significant? In what way? and what does this mean
    for all of us?

Two questions above are heavily related to our current concern: climate
change. To answer both of the above questions, we choose the dataset
taken from WorldBank API, which is CRU-x0.5, for the precipitation and
surface temperature of each country annually from 1901 to 2022. Due to
this being a model approach to data (some entries not collected by
devices at that time), there could be some inaccuracies in the data.
However, CRU-x0.5 have most of its data taken by devices and real
measurements, so it is a negligible problem.

There are two tables in our dataset, surface_temp and precipitation.

-   **surface_temp**:

    -   **code/ISO_A3 (string):** The ISO_A3 code of each country in the
        world. Some country may not have this code due to not being
        recognized globally, so they are not present in this dataset
        (for example, Somaliland).

    -   **country (string):** Name of the country.

    -   **1901-07 -\> 2022-07 (float)**: The surface temperature of each
        year, from 1901 to 2022, collected in 7 a.m. at each morning.
        Each column is a mean of surface temperature at that year.

-   **precipitation**:

    -   **code/ISO_A3 (string):** The ISO_A3 code of each country in the
        world. Some country may not have this code due to not being
        recognized globally, so they are not present in this dataset
        (for example, Somaliland).

    -   **country (string):** Name of the country

    -   **1901-07 -\> 2022-07 (float)**: The mean precipitation of each
        year, from 1901 to 2022, collected in 7 a.m. at each morning.
        Each column is a total precipitation at that year.

## Justification of approach

Static charts struggle to depict how temperature and precipitation
fluctuate across large areas and time periods. An interactive map
addresses this issue by employing color shifts to indicate trends and
geographical variances. Users can also examine individual years and zoom
in for a more in-depth view, making it an excellent tool for
comprehending complicated climate data.

### Tendency of the global surface temperature/precipitation changes

To have a comparison and intuitively see the trend of surface
temperature/ precipitation, visualizing the differences among years
would be more effective than separately showing the statistics of each
year. Indeed, an animated global map that indicates how the temperature
in each country has increased/ decreased over the years compared to that
statistic in 1901 is necessary. The reason is that the differences can
be emphasized thanks to modifications in colors and their intensities
while the structured global map remains unchanged.  We use 2 different
opposite, highly attractive colors: blues: for cooler areas and red: for
hotter ones to help people realize the global trends of specific surface
temperatures just by quickly scanning the map.

Moreover, because of the large dataset including temperature changes in
more than a century, the effort to slowly visualize the map in each year
would be extremely time-consuming. Therefore, we decided to choose only
a 50ms duration for each frame during the animation for better data
visualization experiences. However, the fast refresh animation thus has
to trade off the loss in the information provided by each frame.
Specifically,  users might not have enough time to get essential
information ( global trends, specific information about certain areas)
or access the global maps in arbitrary years.  Therefore, we decided the
add another slider control map to allow users to visit each specific
year as well as zoom in/out to research information about certain areas.

Averaging the temperature/precipitation data for countries in a
continent/ region and allowing users to observe the changes at the
regional level bringing back a more general and systematic perspective.

### 3D Interactive Globe Visualization

While a simple 2D map is adequate, it sometimes struggle to fully
present all the data and insights in the dataset. In this case, an
interactive 3D visualization is an improvement, because not only a 3D
visualization will impress the audience at the beginning, but it can
also present complex concepts such as accurate geographical location in
the visualization. Thus, we think that a 3D interactive globe
visualization would be ideal in this case.

## Code

```{r}
# CRAN version
#install.packages('leaflet')
#install.packages("sf")
# Or Github version
#if (!require('devtools')) install.packages('devtools')
#devtools::install_github('rstudio/leaflet')

#install.packages("animation")
```

```{r}
df_surface_temp <- read.csv('~/Data_Visualization/COMP4010_Project2/data/temp_change.csv')
df_precipitation <- read.csv('~/Data_Visualization/COMP4010_Project2/data/precipitation_change.csv')
```

```{r}
# Load and read the GeoJSON data
library(sf)
countries <- st_read("~/Data_Visualization/COMP4010_Project2/data/ne_110m_admin_0_countries.geojson")
```

```{r}
library(sf)
library(leaflet)
library(RColorBrewer)

# Assume df_precipitation is already read and available
column_name <- "X1993.07"

# Merging country data with precipitation data
merged_data <- merge(countries, df_precipitation, by.x = "ISO_A3", by.y = "code")

# Check the head of the merged data to ensure merge was successful
head(merged_data)

# Remove or replace NA values in the precipitation data
merged_data[[column_name]] <- ifelse(is.na(merged_data[[column_name]]), 0, merged_data[[column_name]])

# Creating a color palette
color_pal <- colorQuantile("YlOrRd", merged_data[[column_name]], n = 10)

# Building the map
map <- leaflet(merged_data) %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addPolygons(fillColor = ~color_pal(get(column_name)),
              fillOpacity = 0.8, 
              color = "white", 
              weight = 0.5,
              popup = ~paste(NAME, ":", get(column_name), "mm")) %>%
  addLegend(pal = color_pal, values = merged_data[[column_name]],
            title = "Precipitation (mm)",
            opacity = 0.7,
            position = "bottomright",
            labFormat = labelFormat())

# Print the map
print(map)
```

```{r}
library(shiny)
ui <- fluidPage(
  titlePanel("Global Precipitation from 1901 to 2022"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("year", "Select Year:",
                  min = 1901, max = 2022,
                  value = 1901, step = 1,
                  animate = animationOptions(interval = 1000, loop = TRUE))
    ),
    mainPanel(
      leafletOutput("map")
    )
  )
)
```

```{r}
server <- function(input, output, session) {
  
  # Observe the year input and update the map
  observe({
    column_name <- paste0("X", input$year, ".07")
    
    # Ensure column exists
    if (column_name %in% colnames(df_precipitation)) {
      # Merging country data with precipitation data
      merged_data <- merge(countries, df_precipitation, by.x = "ISO_A3", by.y = "code")
      
      # Remove NA values in the precipitation data
      merged_data[[column_name]] <- ifelse(is.na(merged_data[[column_name]]), 0, merged_data[[column_name]])
      
      # Creating a color palette
      color_pal <- colorQuantile("YlOrRd", merged_data[[column_name]], n = 10)
      
      # Render the map
      output$map <- renderLeaflet({
        leaflet(merged_data) %>%
          addProviderTiles(providers$OpenStreetMap) %>%
          addPolygons(fillColor = ~color_pal(get(column_name)),
                      fillOpacity = 0.8, 
                      color = "white", 
                      weight = 0.5,
                      popup = ~paste(NAME, ":", get(column_name), "mm")) %>%
          addLegend(pal = color_pal, values = merged_data[[column_name]],
                    title = "Precipitation (mm)",
                    opacity = 0.7,
                    position = "bottomright",
                    labFormat = labelFormat())
      })
    }
  })
}

shinyApp(ui, server)
```

```{r}
library(shiny)
library(leaflet)
library(sf)
library(RColorBrewer)

# Define the UI for the application
ui <- fluidPage(
  titlePanel("Global Surface Temperature Change from 1901 to 2022"),
  sidebarLayout(
    sidebarPanel(
      # Slider input for selecting the year
      sliderInput("yearInput", "Select Year",
                  min = 1902, max = 2022, value = 1902, step = 1),
      width = 3
    ),
    mainPanel(
      leafletOutput("map", width = "100%", height = "800px")
    )
  )
)

# Define server logic required to draw the map
server <- function(input, output) {

  output$map <- renderLeaflet({
    # Dynamic column name based on slider input
    column_name <- paste0("X", input$yearInput, ".07")  # Adjust formatting

    # Filter and prepare the data
    data_filtered <- df_surface_temp[, c("code", column_name)]
    names(data_filtered)[names(data_filtered) == column_name] <- 'value'
    merged_data <- merge(countries, data_filtered, by.x = "ISO_A3", by.y = "code")
    merged_data$value <- as.numeric(as.character(merged_data$value))
    merged_data <- merged_data[!is.na(merged_data$value), ]

    # Custom color and opacity function
    getColor <- function(value) {
      if (value == 0) {
        return("yellow")
      } else if (value > 0) {
        red_intensity <- min(0.8, value / max(merged_data$value)+0.2) 

        return(rgb(1, 0, 0, red_intensity))  # Blue with intensity based on value
      } else {
        blue_intensity <- min(0.8,  value / min(merged_data$value)+0.2)

        return(rgb(0, 0, 1, blue_intensity))  # Red with intensity based on value
      }
    }

    # Apply the color function to each value
    merged_data$color <- sapply(merged_data$value, getColor)

    # Create and return the Leaflet map
    leaflet(merged_data) %>%
      addTiles() %>%
      addPolygons(fillColor = ~color,
                  fillOpacity = 0.8,
                  color = "white",
                  weight = 0.5,
                  popup = ~paste(NAME, ":", round(value, 2), "degree C")) %>%
      addLegend(position = "bottomright",
                title = "Temperature (C)",
                opacity = 0.7,
                labFormat = labelFormat(suffix = " C"),
                colors = c("yellow", rgb(1, 0, 0, 0.8), rgb(0, 0, 1, 0.8)),
                labels = c("Non change", "Increase", "Decrease"))
  })
}

# Run the application 
shinyApp(ui = ui, server = server)
```

### 2D Animated Globe Visualization

Initially,  we preprocessed the raw structured data to obtain the
differences in temperature/precipitation in countries from 1902-2022
compared to the based value in 1901.

First, we create an interactive 2D map with a slider to show changes in
precipitation and temperature over the years. When you hover over a
specific country, the map will display its precipitation or temperature
values. These values will be shown in different colors to indicate
whether they are higher than those of other countries or have increased
compared to the previous year.

Then, we generate an animation illustrating changes in temperature/
precipitation in countries based on a shiny app that includes an UI and
a logical server. Specifically:

-   We initialize a Leaflet map with the data, then add polygons,
    dynamic color, and legend onto the generated map

-   Reactive timer and values will be used for iteratively updating a
    new “Year” variable after each 50ms. Therefore, create the animation
    for the map.

-   The getColor() function is responsible for extracting the proper
    color for appropriate differences compared to the data in 1901 (the
    more different, the higher the color intensity) 

```{r}
library(shiny)
library(leaflet)
library(animation)

server <- function(input, output, session) {
  # Reactive timer
  timer <- reactiveTimer(50)  # Update every 1 second
  
  # Create a reactive value to store the current year
  current_year <- reactiveVal(1901)
  
  # Update the current year every second
  observeEvent(timer(), {
    new_year <- current_year() + 1
    if (new_year > 2022) {
      new_year <- 1901  # Reset to 1901 after reaching 2022
    }
    current_year(new_year)
  })
  
  output$map <- renderLeaflet({
    # Dynamic column name based on current year
    column_name <- paste0("X", current_year(), ".07")  # Adjust formatting if necessary
    # Filter and prepare the data
    data_filtered <- df_surface_temp[, c("code", column_name)]
    names(data_filtered)[names(data_filtered) == column_name] <- 'value'
    merged_data <- merge(countries, data_filtered, by.x = "ISO_A3", by.y = "code")
    merged_data$value <- as.numeric(as.character(merged_data$value))
    merged_data <- merged_data[!is.na(merged_data$value), ]
    minv <- min(merged_data$value)
    maxv <- max(merged_data$value)

    # Custom color and opacity function
    getColor <- function(value) {
      if (value == 0) {
        return("yellow")
      } else if (value > 0) {
        red_intensity <- min(0.8, value / max(merged_data$value) + 0.2)
        return(rgb(1, 0, 0, red_intensity))  # Red with intensity based on value
      } else {
        blue_intensity <- min(0.8, value / min(merged_data$value) + 0.2)
        return(rgb(0, 0, 1, blue_intensity))  # Blue with intensity based on value
      }
    }
     # Define the color palette for the legend
    pal <- colorNumeric(
      palette = c(rgb(0, 0, 1, 1), "yellow", rgb(1, 0, 0, 1)),
      domain = c(minv, maxv)
    )

    # Apply the color function to each value
    merged_data$color <- sapply(merged_data$value, getColor)
    # Create and return the Leaflet map
    leaflet(merged_data) %>%
      addTiles() %>%
      addPolygons(fillColor = ~color,
                  fillOpacity = 0.8,
                  color = "white",
                  weight = 0.5,
                  popup = ~paste(NAME, ":", round(value, 2), "degree C")) %>%
      addControl(HTML(paste0("<b>Year: ", current_year(), "</b>")), position = "topleft")%>%
      addLegend(position = "bottomright",
                pal = pal,
                values = merged_data$value,
                title = "Temperature Change (degree C)",
                opacity = 0.7,
                 labFormat = labelFormat())
  })

  # Render the animation
  output$animation <- renderUI({
    animation::saveGIF({
      for (year in 1901:2022) {
        current_year(year)
        map_frame <- isolate(output$map())
        print(map_frame)
      }
    }, movie.name = "temperature_animation.gif")
    tags$img(src = "temperature_animation.gif", width = "100%")
  })
}

ui <- fluidPage(
  titlePanel("Temperature change compared to 1901"),
  mainPanel(
    leafletOutput("map", width = "100%",height = "800px"),
    uiOutput("animation")
  )
)

shinyApp(ui = ui, server = server)
```

### 3D Interactive Globe Visualization

In this part, we have a 3D Globe rendered using JavaScript.

It can be access by running a live server on the machine and open the
respective HTML file. There are two globes in our project: The globe
about temperature change and precipitation change compared to Year 1901.
The functionalities are:

-   Globe rotation: The globe can be rotated by dragging the sphere
    around.

-   Country hovering and highlight: A country would be highlighted and
    the information about that country is also displayed.

-   Year Slider: The user can use the slider to choose which year will
    be displayed on the globe.

-   Legend: A gradient legend will be displayed in the bottom left of
    the page.

## Discussion

### Changes in Temperature compared to 1901.

As we observe from the animation, the initial trend of global warming
was not unilateral because some areas were hotter while others
experienced a decrease in surface temperature. However, in general,
countries in the same region ( North America, Southeast Asia, Europe,
etc.) tend to have similar changes during certain periods. For instance,
from the 1920s to 1940s, Asian countries ( Russia, China, etc.) observed
a slight increase in their temperature while the USA and Canada both saw
a drop in temperature compared to 1901. Overall, industrialization was
still in the first steps and the global population gradually decreased
due to some wars from the 1920s - to the 1950s, the variance/
differences in temperature were not significant ( mostly below 2 degrees
C).

The changes have become more dramatic and united since the explosion of
the computer/ digital industry in the 1990s. From that point, the
variance of surface temperature increases up to 3.5 degrees C in some
countries ( Russia, Japan, Mongolia ). Moreover, since the 1990s, we can
also observe more red areas on the map, which indicates the significant
increase in global temperature. The number of blue areas notably dropped
and reached its lowest record in 2015 with only 3 ( over \~ 200) 
countries, which are Liberia, Chile, and New Caledonia, having lower
temperatures compared to 1901. Even though surface temperatures in some
countries have slightly decreased in the 2020s thanks to several timely
strategies, the overall status of global warming is in a warning state.

### Changes in Precipitation compared to 1901.

Although the temperature has significantly fluctuated through the years
from 1901 - 2022, we can observe a much more stable trend in
precipitation. Overall, regions having high precipitation, South America
and Southeast Asia in particular, remain in their positions in the rank
of global precipitation by maintaining the average annual rain level
from 1800-2800 mm. For other countries, the precipitation has a slight
increase ( 40 -50 mm in annual indicator) in most of them, for instance,
China (550mm in 1901 to 600mm in 2022) or the USA ( 650mm in 1901 to 704
mm in 2022). However, there is still a minority of regions/countries
where people experience a drop in annual precipitation, for instance,
Mongolia ( 224mm in 1901 to 190mm in 2022).

### Linear Regression

The linear regression analysis of surface temperature/precipiation
changes for China, the United Kingdom, Russia, the United States, and
Vietnam reveals a significant upward trend in temperatures from the
early 1900s to 2022. This consistent increase across multiple regions
highlights the pervasive impact of global climate change, driven by
factors such as industrialization, urbanization, and increased
greenhouse gas emissions. For instance, China shows a clear upward trend
in temperatures from the early 1900s to 2022, with significant
fluctuations, particularly in recent decades. This increase is likely
due to rapid industrialization and urbanization.

Focusing on Vietnam, the data shows a more gradual but steady rise in
surface temperatures. This trend is less pronounced compared to some
other countries but still indicates the effects of global warming. The
increasing temperatures in Vietnam can impact agriculture, water
resources, and local ecosystems, necessitating adaptive measures to
mitigate adverse effects.

```{r}
df_temp_change <- read.csv('~/Data_Visualization/COMP4010_Project2/data/surfaces_temp.csv')
#df_temp_vn <- df_temp_change[df_temp_change$code == 'VNM',]

# Filter data for specific countries
#countries <- c("United States of America", "United Kingdom", "Russian Federation", "China") 
countries_code <- c("USA", "RUS", "GBR", "CHN", "VNM") 
# ISO country codes for USA, UK, Russia, China
df_countries <- df_temp_change[df_temp_change$code %in% countries_code,]

# View the filtered data
head(df_countries)
```

```{r}
library(reshape2)

# Reshape the data to long format
df_long <- melt(df_countries, id.vars = c("code", "name"))

# Extract year from variable names
df_long$year <- as.numeric(sub("X(\\d{4}).*", "\\1", df_long$variable))

# Rename columns for clarity
df_long <- df_long[, c("code", "name", "year", "value")]
colnames(df_long) <- c("Code", "Name", "Year", "Value")

# View the reshaped data
head(df_long)
```

```{r}
library(ggplot2)

# Perform linear regression and plot for each country
unique_countries <- unique(df_long$Code)

# Create a plot for each country
for (country in unique_countries) {
  country_data <- df_long[df_long$Code == country, ]
  
  # Perform Linear Regression
  model <- lm(Value ~ Year, data = country_data)
  summary(model)
  
  # Plot the Data with Regression Line
  p <- ggplot(country_data, aes(x = Year, y = Value)) +
    geom_point() +
    geom_line() +
    geom_smooth(method = "lm", col = "blue") +
    labs(title = paste("Linear Regression of Surface Temperature Over Years for", unique(country_data$Name)),
         x = "Year",
         y = "Temperature") +
    theme_minimal()
  
  print(p)
}
```

```{r}
# Predict the value for the next year (2023) for each country
for (country in unique_countries) {
  country_data <- df_long[df_long$Code == country, ]
  model <- lm(Value ~ Year, data = country_data)
  new_data <- data.frame(Year = 2025)
  prediction <- predict(model, newdata = new_data)
  print(paste("Predicted value for the year 2025 for", unique(country_data$Name), ":", prediction))
}
```

```{r}
df_precip <- read.csv('~/Data_Visualization/COMP4010_Project2/data/precipitation.csv')
#df_temp_vn <- df_temp_change[df_temp_change$code == 'VNM',]

# Filter data for specific countries
#countries <- c("United States of America", "United Kingdom", "Russian Federation", "China") 
countries_code_pre <- c("USA", "RUS", "GBR", "CHN", "VNM") 
# ISO country codes for USA, UK, Russia, China
df_countries_pre <- df_precip[df_precip$code %in% countries_code_pre,]

# View the filtered data
head(df_countries_pre)
```

```{r}
library(reshape2)

# Reshape the data to long format
df_long_pre <- melt(df_countries_pre, id.vars = c("code", "name"))

# Extract year from variable names
df_long_pre$year <- as.numeric(sub("X(\\d{4}).*", "\\1", df_long_pre$variable))

# Rename columns for clarity
df_long_pre <- df_long_pre[, c("code", "name", "year", "value")]
colnames(df_long_pre) <- c("Code", "Name", "Year", "Value")

# View the reshaped data
head(df_long_pre)
```

```{r}
library(ggplot2)

# Perform linear regression and plot for each country
unique_countries_pre <- unique(df_long_pre$Code)

# Create a plot for each country
for (country in unique_countries_pre) {
  country_data <- df_long_pre[df_long_pre$Code == country, ]
  
  # Perform Linear Regression
  model <- lm(Value ~ Year, data = country_data)
  summary(model)
  
  # Plot the Data with Regression Line
  p <- ggplot(country_data, aes(x = Year, y = Value)) +
    geom_point() +
    geom_line() +
    geom_smooth(method = "lm", col = "blue") +
    labs(title = paste("Linear Regression of Precipitation Over Years for", unique(country_data$Name)),
         x = "Year",
         y = "Precipitation (mm)") +
    theme_minimal()
  
  print(p)
}
```

```{r}
# Predict the value for the next year (2023) for each country
for (country in unique_countries) {
  country_data <- df_long_pre[df_long_pre$Code == country, ]
  model <- lm(Value ~ Year, data = country_data)
  new_data <- data.frame(Year = 2025)
  prediction <- predict(model, newdata = new_data)
  print(paste("Predicted value for the year 2025 for", unique(country_data$Name), ":", prediction))
}
```

## Limitations

The relationship between temperature and precipitation is complex and
often non-linear. While warmer air can hold more moisture, leading to
potentially heavier precipitation events, there are tipping points.
Extremely high temperatures can suppress precipitation formation, and
regional atmospheric circulation patterns significantly influence
rainfall distribution.  Understanding these non-linear dynamics is
crucial for accurate weather prediction and climate change modeling.

Additionally, ensuring the accuracy and reliability of imputed data
poses its own set of challenges. Despite these limitations, the
application remains a powerful tool for visualizing climate data where
available.

## Future directions

We propose creating an interactive map application with 2D and 3D views
to visualize precipitation and temperature data dynamically. The app
will update daily and provide notifications for significant changes.
Users can explore detailed maps, viewing temperature and precipitation
data down to the city level. This will be achieved using Shiny for the
web interface, Leaflet for 2D maps, and CesiumJS for 3D visualizations.
A robust backend will handle data updates and alert mechanisms. This
tool aims to provide an accessible and detailed view of climate data,
supporting better decision-making and awareness of environmental
changes. Also, the 3D visualization is still lacking in terms of clarity
and integration. In the future, we intend to incorporate the plot of
temperature over year for each of the country when you clicked on it, as
well as the linear regression for better interpretability.
